Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-07-31 joint_inf.tex
update plot

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Figure~\ref{fig:farris-fels-top}.
We parameterize the branches of these trees \textbf{not with the standard notion of branch length in terms of number of substitutions per site}, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$.
This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform.

@@

Consider only the Farris-tree topology.
Assume we know the true values of $x'$ and $y'$ but not $w'$.
In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
In this case, the likelihood as a function of $w'$ is proportional to either
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + \log((1+w'))
\end{align*}
or
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + (1 - p_{13})\cdot\log((1+w')) \\
    &\qquad + p_{13}\cdot\log((1-w')).
\end{align*}
When computing the global minimum for $0 \le w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}$ in Figure~\ref{fig:w_hat}.
The estimate for $\hat{w}$ will typically be an overestimate, and in fact is almost always identically equal to 1, indicating no change along that particular branch.

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{w_hat_heatmap}
\caption{
    Estimating $\hat{w}$ for $x'$ and $y'$ fixed at their true values.
}
\label{fig:w_hat}
\end{figure}

Since we do not know the true values of $x$ and $y$, standard methods will not obtain the situation in Figure~\ref{fig:w_hat}.
The form of the likelihood will alter the fitted values of $x$ and $y$ so that even with perfect information we will not necessarily obtain the true parameters back.
If we estimate $(x', y', w')$ jointly using the same approach above, we obtain the fit $\hat{w}$ in Figure~\ref{fig:w_hat_joint}.

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{w_hat_heatmap_joint}
\caption{
    Estimating $(\hat{x}, \hat{y}, \hat{w})$.
}
\label{fig:w_hat}
\end{figure}

We no longer consistently overestimate $\hat{w}$, but the estimates are surprisingly unstable, with estimates of 1 for $\hat{w}$ scattered throughout the $(x,y)$ lattice.

>>>>

Define the Farris topology $\tau_1$ and the Felsenstein topology $\tau_2$ as in Fig.~\ref{fig:farris-fels-top}.
We parameterize the branches of these trees \textbf{not with the standard notion of branch length in terms of number of substitutions per site}, but with a parameter such that the probability of a substitution on a branch with parameter $\theta$ is $p_\theta = (1-\theta)/2$ while the probability of no substitution is $(1+\theta)/2$.
This parametrization has been called edge ``fidelity'' as it quantifies the fidelity of transmission of the ancestral state along an edge \cite{Matsen2007-jq}.
Fidelities have useful algebraic properties, such as the ability to write generating probabilities using the Hadamard transform.

@@

Consider only the Farris-tree topology.
Assume we know the true values of $x'$ and $y'$ but not $w'$.
In other words, we fix $x'=x$ and $y'=y$ to their true, generating values and wish to estimate the internal branch parameter $w'$.
In this case, the likelihood as a function of $w'$ is proportional to either
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + \log((1+w'))
\end{align*}
or
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, w'; \fullAncestralSplitPartitions(\tau_1,w'))
    &\propto        p_{\emptyset}  \cdot\log(1+x^2+y^2+4xyw'+x^2y^2) \\
    &\qquad + p_{13}         \cdot\log(1+x^2+y^2-4xyw'+x^2y^2) \\
    &\qquad + (1 - p_{13})\cdot\log((1+w')) \\
    &\qquad + p_{13}\cdot\log((1-w')).
\end{align*}
When computing the global minimum for $0 \le w' \le 1$ using basin-hopping, we obtain the fit $\hat{w}$ in Fig.~\ref{fig:w_hat}.
The estimate for $\hat{w}$ will typically be an overestimate, and in fact is almost always identically equal to 1, indicating no change along that particular branch.

\begin{figure}
\centering
    \begin{tabular}{cc}
\includegraphics[width=.45\textwidth]{w_hat_heatmap} &
\includegraphics[width=.45\textwidth]{w_hat_heatmap_no_joint}\\
Joint inference & Marginal inference \\
    \end{tabular}
\caption{
    Estimating $\hat{w}$ for $x'$ and $y'$ fixed at their true values.
}
\label{fig:w_hat}
\end{figure}

Since we do not know the true values of $x$ and $y$, standard methods will not be in the situation in Figure~\ref{fig:w_hat}.
The form of the likelihood will alter the fitted values of $x$ and $y$ so that even with perfect information we will not necessarily estimate $\hat{x}=x$ and $\hat{y}=y$.
If we estimate $(x', y', w')$ jointly using the same approach above, we obtain the fit $\hat{w}$ in Figure~\ref{fig:w_hat_joint}.

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{w_hat_heatmap_joint}
\caption{
    Estimating $(\hat{x}, \hat{y}, \hat{w})$.
}
\label{fig:w_hat_joint}
\end{figure}

We no longer consistently overestimate $\hat{w}$, but the estimates are surprisingly unstable, with estimates of 1 for $\hat{w}$ scattered throughout the $(x,y)$ lattice.
Even given the correct Farris topology we obtain poor estimates for the value of the interior branch length.
